Learning communication systems through recursive pragmatic reasoning about other agents

### 2013-05-31 praglearn.tex
first pass at conclusion

Even before they are proficient language users, children learn the
meanings of words from their pattern of use in context. For example,
in demonstrations of the ``disambiguation effect,'' a novel word
(e.g. ``dax'') is used in a context containing both a novel and a
familiar object. From quite young, children tend to make the inference
that ``dax'' refers to the novel object \cite{markman1988}. In another
demonstration, even young toddlers can extract mappings between words
and objects from a pattern of individually-ambiguous co-occurrences
between pairs of words and pairs of objects \cite{smith2008}. Taking
inspiration from children's learning prowess, recent work in the field
of grounded language learning has explored the possibility that even
relatively complex word meanings can be inferred from corpus data
(e.g. \cite{zettlemoyer2005,chen2008,frank2009,kwiatkowski2010,johnson2012}).

Recursive pragmatic inferences such as those described above could be
an important part of language learning.\footnote{Very young children
  make inferences that are often labeled as ``pragmatic'' in that they
  involve reasoning about context \cite{clark1988,baldwin1993}, though
  they also show weaknesses in specific inferences (e.g. strengthening
  ``some'' to ``some but not all'' \cite{papafragou2003}). Here we
  remain agnostic about the age at which children are able to make
  such inferences robustly, as it may vary depending on the linguistic
  materials being used in the inference \cite{barner2011}.} For
example, the disambiguation inference described above can emerge from
such reasoning (``why would the speaker have said ``dax'' if there was
a word for the known object that she knew I would know'')
\cite{clark1988}. In addition, there are a variety of intriguing cases
in which listeners seem to create situation- and task-specific ways of
referring to particular objects. For example, when asked to refer to
idiosyncratic geometric shapes, over the course of an experimental
session, participants create conventionalized descriptions that allow
them to perform accurately even though they do not begin with a shared
label \cite{krauss1964,clark1986}. Yet models of grounded language
learning have failed to take pragmatic factors into account.

@@

\begin{table}[t]
\label{tab:results}
\begin{center}
\begin{tabular}{lccccc}
\toprule
Phenomenon & Ref. & WL & PI & PI+U & PIL \\
\midrule
Interpreting specificity implicature & \cite{grice1975} &  & x &x & x\\
Interpreting Horn implicature & \cite{horn1984} & &  & x& x\\
Cross-situational learning & \cite{smith2008} & x& & & x \\
Disambiguating new words using old words & \cite{markman1988} &x  & & & x \\
Disambiguation without mapping & \cite{horst2008} &  & & & x \\
Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
Lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
\hline
\end{tabular}
\end{center}
\caption{Empirical results and references. WL refers to the word
  learning model of \cite{frank2009}; PI refers to the recursive
  pragmatic inference model of \cite{frank2012}; PI+U refers to the
  pragmatic inference model of \cite{bergen2012} which includes
  lexical uncertainty, marginalizes it out, and then recurses. Our
  current model is referred to here as PIL, or pragmatic inference and
  learning.}
\end{table}

@@

Children, when presented with both a novel and a familiar object
(e.g. an eggbeater and a ball), will treat a novel label
(e.g. ``dax'') as referring to the novel object, for example by
supplying the eggbeater when asked to ``give me the dax''
\cite{markman1988}. This phenomenon is sometimes referred to as
``mutual exclusivity'' or even ``fast mapping.'' Simple probabilistic word learning models can produce a
similar pattern of findings \cite{frank2009}, but all such models assume that learners retain the
mapping between novel word and novel object demonstrated in the
experimental situation. This observation is contradicted, however, by
evidence that children often do not retain the mappings that are
demonstrated by their inferences in the moment \cite{horst2008}.

Our model provides an intriguing possible explanation of this finding:
given a single disambiguation situation, the model gives a substantial
probability (e.g. 75\%) that the speaker is referring to the novel
object. Nevertheless, this inference is not accompanied by an increased belief
that the novel word literally refers to this object. The learner's
interpretation arises not from lexical mapping but instead from a kind
of 
specificity implicature: the listener knows that the familiar word
\emph{does not} refer to the novel object---hence the novel word (even
if it refers to the familiar object as well) is the best way to refer
to the novel object. Nevertheless, on repeated exposure to the same
novel word, novel object situation, the learner does learn the mapping
as part of the lexicon (congruent with other data on repeated training
on disambiguation situations \cite{bion2012}).

@@

@@

\subsection{Emergence of efficient communicative conventions}

@@

Something
fundamentally different happens in this case than in the specificity
implicature case (\S\ref{sec:learning-specificity-implic}), however. There, we
saw that our model could hear pragmatically strengthened usages
of words while recovering the underlying literal meaning. In the Horn
case, though, the pragmatic strengthening depends essentially on
ignorance about literal meaning. As the agents gather more data,
the implicature disappears, to be replaced by a belief that
``cheap'' \textit{literally} means \textsc{common}. To demonstrate
this phenomenon, we queried each agent in each dyad about how they would refer to
or interpret each object and word, \textit{if} the two objects were
equally common, which cancels the Horn implicature. As shown in
Fig.~\ref{fig:emergence-average} (right), after 30 turns, in nearly 70\%
of dyads both $S$ and $L$ believed in the ``good'' mapping even in
this implicature-free case, while less than 20\% believed in the
``bad'' mapping (with the rest being inconsistent).

@@

If each agent in a multi-agent setting assumes the existence of a
shared, conventional norm underlying everyone's behaviour, and further
assumes that other agents are knowledgeable about this norm while they
themselves are not, then we are able to preserve most of the normative
Bayesian framework, explicitly model both learning and recursive
social reasoning simultaneously within the same framework, and avoid
paradox. The resulting theory unifies two previously separate
strands of modelling work on language usage and acquisition,
accounting for a superset of previous phenomena, while producing a
variety of new qualitative patterns. In particular, we make new

the deviations from rationality -- assumption that words have
meanings, suspicion that everyone else knows what they are -- are
highly interpretable and intuitively compelling, and while like any
deviations from rationality they necessarily introduce biases, these
biases appear in this case to improve the global performance of the
system.

if you're trying to collaborate with an agent who soft-maximizes
utility, but there are multiple possible rules which might underlie
their utility function, you should assign more weight to those rules
which provide a sharp distinction in utility for different actions --
which will generally be the rules that allow for higher expected
utility. this actually explains both the uniform prior and the horn
implicature emergence results.

provides new insight into use of pragmatic reasoning in word learning,
the learning of literal meanings given pragmatically strengthened
input, and makes the interesting prediction that some implicatures are
intrinsically more stable, while others if consistently supported will
tend to enter the lexicon and become part of the literal meaning of
the relevant words.

zipf, piantidosi et al

>>>>

Even before they are proficient language users, children learn the
meanings of words from their pattern of use in context. For example,
in demonstrations of the ``disambiguation effect,'' a novel word
(e.g.~``dax'') is used in a context containing both a novel and a
familiar object. From quite young, children tend to make the inference
that ``dax'' refers to the novel object \cite{markman1988}. In another
demonstration, even young toddlers can extract mappings between words
and objects from a pattern of individually-ambiguous co-occurrences
between pairs of words and pairs of objects \cite{smith2008}. Taking
inspiration from children's learning prowess, recent work in the field
of grounded language learning has explored the possibility that even
relatively complex word meanings can be inferred from corpus data
(e.g.~\cite{zettlemoyer2005,chen2008,frank2009,kwiatkowski2010,johnson2012}).

Recursive pragmatic inferences such as those described above could be
an important part of language learning.\footnote{Very young children
  make inferences that are often labeled as ``pragmatic'' in that they
  involve reasoning about context \cite{clark1988,baldwin1993}, though
  they also show weaknesses in specific inferences (e.g.~strengthening
  ``some'' to ``some but not all'' \cite{papafragou2003}). Here we
  remain agnostic about the age at which children are able to make
  such inferences robustly, as it may vary depending on the linguistic
  materials being used in the inference \cite{barner2011}.} For
example, the disambiguation inference described above can emerge from
such reasoning (``why would the speaker have said ``dax'' if there was
a word for the known object that she knew I would know'')
\cite{clark1988}. In addition, there are a variety of intriguing cases
in which listeners seem to create situation- and task-specific ways of
referring to particular objects. For example, when asked to refer to
idiosyncratic geometric shapes, over the course of an experimental
session, participants create conventionalized descriptions that allow
them to perform accurately even though they do not begin with a shared
label \cite{krauss1964,clark1986}. Yet models of grounded language
learning have failed to take pragmatic factors into account.

@@

\begin{table}[t]
\label{tab:results}
\begin{center}
\begin{tabular}{lccccc}
\toprule
Phenomenon & Ref. & WL & PI & PI+U & PI+WL \\
\midrule
Interpreting specificity implicature & \cite{grice1975} &  & x &x & x\\
Interpreting Horn implicature & \cite{horn1984} & &  & x& x\\
Cross-situational learning & \cite{smith2008} & x& & & x \\
Disambiguating new words using old words & \cite{markman1988} &x  & & & x \\
Disambiguation without mapping & \cite{horst2008} &  & & & x \\
Learning/lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
\hline
\end{tabular}
\end{center}
\caption{Empirical results and references. WL refers to the word
  learning model of \cite{frank2009}; PI refers to the recursive
  pragmatic inference model of \cite{frank2012}; PI+U refers to the
  pragmatic inference model of \cite{bergen2012} which includes
  lexical uncertainty, marginalizes it out, and then recurses. Our
  current model is referred to here as PI+WL, and combines pragmatic
  inference with word learning.}
\end{table}

@@

Children, when presented with both a novel and a familiar object
(e.g.~an eggbeater and a ball), will treat a novel label
(e.g.~``dax'') as referring to the novel object, for example by
supplying the eggbeater when asked to ``give me the dax''
\cite{markman1988}. This phenomenon is sometimes referred to as
``mutual exclusivity'' or even ``fast mapping.'' Simple probabilistic word learning models can produce a
similar pattern of findings \cite{frank2009}, but all such models assume that learners retain the
mapping between novel word and novel object demonstrated in the
experimental situation. This observation is contradicted, however, by
evidence that children often do not retain the mappings that are
demonstrated by their inferences in the moment \cite{horst2008}.

Our model provides an intriguing possible explanation of this finding:
given a single disambiguation situation, the model gives a substantial
probability (e.g.~75\%) that the speaker is referring to the novel
object. Nevertheless, this inference is not accompanied by an increased belief
that the novel word literally refers to this object. The learner's
interpretation arises not from lexical mapping but instead from a kind
of 
specificity implicature: the listener knows that the familiar word
\emph{does not} refer to the novel object---hence the novel word (even
if it refers to the familiar object as well) is the best way to refer
to the novel object. Nevertheless, on repeated exposure to the same
novel word, novel object situation, the learner does learn the mapping
as part of the lexicon (congruent with other data on repeated training
on disambiguation situations \cite{bion2012}).

@@

agents not sure whether some literally means some-not-all or not,
could lexicalize given the right kind of input, but regular pragmatic
strengthening alone is not enough

@@

\subsection{Emergence of efficient communicative conventions}
\label{sec:emergence}

@@

Something
fundamentally different happens in this case than in the specificity
implicature case (\S\ref{sec:learning-specificity-implic}), however. There, we
saw that our model could hear pragmatically strengthened usages
of words while recovering the underlying literal meaning. In the Horn
case, though, the pragmatic strengthening depends essentially on
ignorance about literal meaning. As the agents gather more data,
the implicature disappears, to be replaced by a belief that
``cheap'' \textit{literally} means \textsc{common}. To demonstrate
this phenomenon, we queried each agent in each dyad about how they would refer to
or interpret each object and word, \textit{if} the two objects were
equally common, which cancels the Horn implicature. As shown in
Fig.~\ref{fig:emergence-average} (right), after 30 turns, in nearly 70\%
of dyads both $S$ and $L$ believed in the ``good'' mapping even in
this implicature-free case, while less than 20\% believed in the
``bad'' mapping (with the rest being inconsistent). This makes the
interesting prediction that while specificity implicatures can survive
indefinitely in a language system, any Horn implicatures that arise
consistently should transition to becoming part of the literal meaning
of the words involved.

@@

If each agent in a multi-agent setting assumes the existence of a
shared, conventional norm underlying everyone's behaviour, and further
assumes that other agents are knowledgeable about this norm while they
themselves are not, then we are able to simultaneously preserve most
of the normative Bayesian framework, explicitly model learning and
recursive social reasoning within the same framework, and avoid
paradox. The resulting model unifies two previously separate strands
of modelling work on language usage and acquisition, and -- having
done so -- accounts for not just the phenomena these previous models
accounted for, but a variety of new ones as well. We propose new
explanations of the child language acquisition phenomena of
disambiguation without mapping \cite{horst2008} and the acquisition of
scalar adjectives (XX ask Mike what the levinson cite is really, see
also table), and demonstrate that our model is capable of developing
novel and efficient communicative systems through iterated learning
within the context of a single simulated conversation.

Accomplishing this requires limited deviations from rationality, and
as such necessarily introduces biases in agent behaviour. We've
identified two such biases here: the tendency for pragmatic speakers
and listeners to accentuate useful, sparse patterns in their
communicative systems (\S\ref{sec:emergence}), and for cheap words to
be assigned to common objects (\S\ref{sec:horn-emergence}).
Remarkably, while both are locally sub-optimal for individual agents,
they both systematically drive the overall communicative system
towards greater global efficiency. In the long term, these processes
should leave their mark on the structure of the language itself, which
may provide a partial explanation for how languages become so tightly
optimized for effective communication \cite{zipf1936,piantadosi2011}.

% the deviations from rationality -- assumption that words have
% meanings, suspicion that everyone else knows what they are -- are
% highly interpretable and intuitively compelling, and while like any
% deviations from rationality they necessarily introduce biases, these
% biases appear in this case to improve the global performance of the
% system.

% if you're trying to collaborate with an agent who soft-maximizes
% utility, but there are multiple possible rules which might underlie
% their utility function, you should assign more weight to those rules
% which provide a sharp distinction in utility for different actions --
% which will generally be the rules that allow for higher expected
% utility. this actually explains both the uniform prior and the horn
% implicature emergence results.

% provides new insight into use of pragmatic reasoning in word learning,
% the learning of literal meanings given pragmatically strengthened
% input, and makes the interesting prediction that some implicatures are
% intrinsically more stable, while others if consistently supported will
% tend to enter the lexicon and become part of the literal meaning of
% the relevant words.

% zipf, piantidosi et al
